%
% Copyright 2020, Data61, CSIRO (ABN 41 687 119 230)
%
% SPDX-License-Identifier: CC-BY-SA-4.0
%

\documentclass[11pt,a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage{isabelle,isabellesym}

% this should be the last package used
\usepackage{pdfsetup}

% urls in roman style, theory text in math-similar italics
\urlstyle{rm}
\isabellestyle{tt}


\begin{document}

\title{State Monad Library}
\author{David Greenaway,
  Gerwin Klein,
  Corey Lewis,
  Daniel Matichuk,
  Thomas Sewell}
\maketitle

\begin{abstract}
  This entry contains a library of different state monads with a large set of
  operators, laws, Hoare Logic, weakest precondition rules, and corresponding
  automation. The formalisation is designed for reasoning about total and
  partial correctness, as well as for reasoning about failure separately from
  normal behaviour. Partial correctness in this context ignores program failure.
  Total correctness implies the absence of program failure.

  The main monads formalised in this entry are a non-deterministic state monad
  with failure, and a state-based trace monad for modelling concurrent executions.
  Both support roughly the same set of operators. They come with a standard
  Hoare Logic and Rely-Guarantee logic respectively. The entry also contains an
  option reader monad that combines well with the non-deterministic state monad.
  The option reader monad provides additional operators useful for building
  state projections that can be used both in monadic functions and Hoare-Logic
  assertions, enhancing specification re-use in proofs.

  This monad library is used in the verification of the seL4 microkernel and
  predates some of the monad developments in the Isabelle library. In
  particular, it defines its own syntax for do-notation, which can be overridden
  with the generic monad syntax in the Isabelle library. We have opted not to do
  so by default, because the overloading-based syntax from the Isabelle library
  often necessitates additional type annotations when mixing different monad
  types within one specification. For similar reasons, no attempt is made to
  state generic state monad laws in a type class or locale and then instantiate
  them for the two main monad instances. The resulting duplication from two
  instances is still easy to handle, but if more instances are added to this
  library, additional work on genericity would be useful.

  This library has grown over more than a decade with many substantial
  contributions. We would specifically like to acknowledge the contributions by
  Nelson Billing, Andrew Boyton, Matthew Brecknell, David Cock, Matthias Daum,
  Alejandro Gomez-Londono, Rafal Kolanski, Japheth Lim, Michael McInerney, Toby
  Murray, Lars Noschinski, Edward Pierzchalski, Sean Seefried, Miki Tanaka, Vernon
  Tang, and Simon Windwood.
\end{abstract}

\tableofcontents

\parindent 0pt\parskip 0.5ex

% generated text of all theories
\input{session}

\bibliographystyle{abbrv}
\bibliography{root}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
