%
% Copyright 2014, NICTA
%
% This software may be distributed and modified according to the terms of
% the GNU General Public License version 2. Note that NO WARRANTY is provided.
% See "LICENSE_GPLv2.txt" for details.
%
% @TAG(NICTA_GPL)
%

\chapter{Introduction}
CAmkES is a component platform for embedded microkernel-based systems, offering
many of the standard features available in component platforms.
Some relevant features of CAmkES that are not common to all component platforms
are:
\begin{itemize}
  \item \emph{Explicit composite components.} CAmkES components can be
    assembled to form a re-usable composite, that can then be referenced within
    a containing system.
  \item \emph{Multiple instantiation of a single component.} Multiple copies of
    a component can exist within a system, distinguished by different
    identifiers.
  \item \emph{Multiple implementations of an interface.} A single component can
    implement an interface more than once. This can be useful for providing a
    dedicated interface for each client or functionally different
    implementations.
  \item \emph{Distinction between active and passive components.} Components
    can have a thread of control or be invoked via an event loop. CAmkES
    distinguishes these modes of operation at an architectural level.
  \item \emph{``Provides'' interfaces can be left unsatisfied at runtime.} When
    a component implements a given interface, it does not have to be connected
    to a component that uses that interface. The converse is not true.
  \item \emph{Hardware devices are components.} Hardware devices can be
    specified in ADL that generates interface methods for accessing the device.
\end{itemize}
